Bayesian exponential random graph modeling of KWL Burial network

### 2020-12-08 analysis/paper/paper.Rmd
work on discussion section and run models

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality that is represented by endogenous network effects, low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, gender, ritual activity, and the degree of wealth, to test whether burials with similar attributes tend to form relationships. For example, age-homophily means people of the same age to have the same burial goods. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to control the preferential tendency of formation of relationship. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models. For covariates based on burial attributes, such as age, gender, and our wealth index, we specified a vague prior that follows a normal distribution with mean at 0 and standard deviation at 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 5)) to infer kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity (gwesp) and popularity (gwedegree). We set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on ritual activity (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the priors for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to  model an increased social inequality. This prior information derives from theory about horizontal hierarchies, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for pre-European network and N(-20, 6) for post-European network.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
As an informal way to diagnose model convergence, we first evaluated three diagnostic visual summaries of our MCMC output, including density plots, trace plots, and autocorrelation plots [@Hamra2013], which we include in our Supplementary Online Materials. Then we summarise output from our two models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. We set 10,000 network graphs simulated from the estimated posterior distribution in ERGMs. Our Bayesian GOF diagnostics summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions.This provides a statistical approach to check how well the estimated posterior parameter distribution, based on our hypotheses, can reproduce networks with similar general structural features of the observed networks. We then compared the distribution of our observed networks, the networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network, such as the relative importance of age and sex.

@@

<!--posterior stats for pre-E-->
We compared estimated statistics from the posterior distributions to examine their differences in  structure of the simulated networks (Table XX). For the pre-European network simulations, two nodal covariates, age-homophily and ritual-homophily, have significant effects on the relations between burials. The age homophily shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly, indicating that physical proximity does not affect relations. For the endogenous network effects, both transitivity (gwesp) and popularity (gwdegree) present significant positive effects. The positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. The positive popularity shows a tendency toward centralization that reflects a limited number of burials have many more relations than others. This implies that a few central individuals developed personal networks through wealth accumulation in terms of the flow of goods.

@@

<!-- move this text to another section that is a more logical fit --> 
In addition, since the prior settings will have impacts on the posterior, we compared the posterior estimates with the same prior for all parameters, and those with the informative prior based on our assumptions for some parameters. In general, there are no significant changes in the posterior estimates using different sets of priors. But we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation.

@@

<!--archaeological implication based on the results-->
The covariants for both networks, before and after the European presence, provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:xxx)). We included age, sex, ritual pottery, and a wealth index based on the value of grave goods. The pre-European network shows that burials with pottery as ritual vessels tend to be tied. On the other hand, in the post-European network ties tend to be between burials within the same level of wealth. Moreover, in the post-European network, burials with ritual vessels tend not to be tied. This illustrates that different social and economic mechanisms determined relations between burials for the two periods. Before European arrival, ritual behaviors seem to have been important status indicators, relative to the use of prestige goods. After European presence was established, status indicators shifted from ritual to wealth differences.

Distance between burials did not correlate to the network formation... <!-- anthropological concept here -->

Demographic information, such as age and sex, do not show any correlations with network formation for either period. This may be due to the unavailability of ge and sex values for many burials in our sample.

@@

<!-- fit into a big anthropological picture-->

<!-- methodological implication-->
In this paper, we used burial data as a proxy to explore the social relations based on the argument that burials could be viewed as a social representation [@Binford1971; @Saxe1970]. We also note that the issue of disconnection between the role of the deceased in the burial context and their real life, which can be resolved by examined different evidence out of the burial context [@Chapman2003]. The ERGM network analysis within a Bayesian framework provides a robust way to examine relations between burials based on grave goods and attributes associated with demography and mortuary practices. The adoption of Bayesian inference on ERGM can solve the issues of computational tractability and degeneracy in parameter estimation in ERGM [@Caimo2017]. In addition, a Bayesian framework reduced the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. We demonstrated a plausible and robust method to test the hypothesized model of chronological changes in burial structures with probabilistic interpretation. Although some network statistics of covariants do not show the pattern as we assumed, the Bayesian approach is an efficient and effective method for evaluating the endogenous network effects relative to the overall structure of burials. This structure can be viewed as a representation of a given society.

<!-- more discussion of nodal attributes -->
Age and gender attributes serve as indicators for distinguishing ascribed or achieved status in a society [citation].

>>>>

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality that is represented by endogenous network effects, low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to test whether burials with similar attributes tend to form relationships. For example, age-homophily means people of the same age to have the same burial goods. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to control the preferential tendency of formation of relationship. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models. For covariates based on burial attributes, such as age, sex, and our wealth index, we specified a vague prior that follows a normal distribution with mean at 0 and standard deviation at 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 5)) to infer kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity (gwesp) and popularity (gwedegree). We set the priors to higher transitivity (N(5, 7)), lower popularity (N(4, 7)), and higher covariate effect based on ritual activity (N(2, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the priors for the network after the European arrival to lower transitivity (N(4, 8)), and higher popularity (N(8, 8)) to model an increased social inequality. This prior information derives from theory about horizontal hierarchies, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set differently with N(-17, 7) for pre-European network and N(-20, 6) for post-European network.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
As an informal way to diagnose model convergence, we first evaluated three diagnostic visual summaries of our MCMC output, including density plots, trace plots, and autocorrelation plots [@Hamra2013], which we include in our Supplementary Online Materials. In general, we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation. Then we summarize output from our two models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. We set 10,000 network graphs simulated from the estimated posterior distribution in ERGMs. Our Bayesian GOF diagnostics summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution, based on our hypotheses, can reproduce networks with similar general structural features of the observed networks. We then compared the distribution of our observed networks, the networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network, such as the relative importance of age and sex.

@@

<!--posterior stats for pre-E-->
We compared estimated statistics from the posterior distributions to examine their differences in  structure of the simulated networks (Table XX). For the pre-European network simulations, two nodal covariates, age-homophily and ritual-homophily, have significant effects on the formation of relations between burials. The age homophily shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly, indicating that physical proximity does not affect relations. For the endogenous network effects, both transitivity (gwesp) and popularity (gwdegree) present significant positive effects. The positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. The positive popularity shows a tendency toward centralization that reflects a limited number of burials have many more relations than others. This implies that a few central individuals developed personal networks through wealth accumulation in terms of the flow of goods.

@@

@@

<!--archaeological implication based on the results-->
The covariants for both networks, before and after the European presence, provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:xxx)). We included age, sex, ritual pottery, and a wealth index based on the value of grave goods. The pre-European network shows that burials with pottery as ritual vessels tend to be tied. On the other hand, in the post-European network ties tend to be between burials within the same level of wealth. Moreover, in the post-European network, burials with ritual vessels tend not to be tied. This illustrates that different social and economic mechanisms determined relations between burials for the two periods. Before European arrival, ritual behaviors seem to have been important status indicators, relative to the use of prestige goods. After European presence was established, status indicators shifted from ritual to wealth differences. Distance between burials did not correlate to network formation for both periods that means geographical proximity, implying some kinship association, is not a factor for sharing similar goods between burials. This may suggest that the use of burial goods is more individual-oriented instead of kinship-based sharing. In addition, demographic information, such as age and sex, do not show any correlations with network formation for either period. This may be due to the unavailability of age and sex values for many burials in our sample.

@@

<!-- other evidence and a bigger anthropological picture-->
We used burial data as a proxy to explore the social relations based on the argument that burials could be viewed as a social representation to understand social structures [@Binford1971; @Saxe1970]. We note that there could be an issue of disconnection between the role of the deceased in a burial context and their real life, which should be reduced by examining evidence from the residential context [@Chapman2003]. Previous studies of trade ornaments from the residential area of Kiwulan suggests an uneven spatial distribution since the Spanish and the Dutch were active in northern Taiwan that hints at an increased social differentiation [@LWandBM2020ornament]. This is consistent with our result of a more centralized burial network after the presence of the Europeans, which supports the assumption of social role representation from burials in this case. Since the burial goods used for making ties were treated as prestige goods in local Indigenous culture [@Borao2009; @LiandWu2006], the interrelations indicated by the flow of goods could reflect the process of social changes at Kiwulan in a pericolonial context. After the presence of the Europeans, a more centralized and less cohesive network indicates some individuals were differentiated or privileged to possess more distinctive goods. Moreover, the relationship tends to form between individuals with the same wealth level represented by the rarity of goods. This uneven prestige accumulation might result from differential access to exotic trade relative to the European colonization. The involvement in long-distance exchange often leads to changes in political-economic strategies toward a network mode, especially for small-scale societies, where ambitious individuals are able to build their factions and personal influence through the distribution of high value goods [@Feinman2000].

<!-- methodological implication-->
The ERGM network analysis within a Bayesian framework provides a robust way to examine relations between burials based on grave goods and attributes associated with demography and mortuary practices. The adoption of Bayesian inference on ERGM can solve the issues of computational tractability and degeneracy in parameter estimation in ERGM [@Caimo2017]. In addition, a Bayesian framework reduced the effects of small sample size or missing data commonly present in archaeological data by incorporating prior information and MCMC estimations. We demonstrated a plausible and robust method to test the hypothesized model of chronological changes in burial structures with probabilistic interpretation. Although some network statistics of covariants do not show the pattern as we assumed, the Bayesian approach is an efficient and effective method for evaluating the endogenous network effects relative to the overall structure of burials. This structure can be viewed as a representation of a given society.

<!-- more discussion of nodal attributes -->
Age and sex attributes serve as indicators for distinguishing ascribed or achieved status in a society [citation].
